A Bayesian method for latitudinal temperature gradients in the early Eocene

### 2023-02-07 manuscript/manuscript.Rmd
S4 figure
intro improvements and results on fig s4

---
title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene"

@@

<!-- discarded titles:
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->

@@

Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g. @Hollis2019], and global temperature reconstructions based on those records alone are prone to bias [@Jones2022]. Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps. A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures and precipitation have been created with Gaussian process regression [@Inglis2020; @Chandra2021].

Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. Although some more recent climate models have been able to align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate. The inferred latitudinal gradients depend on the selection of proxies and on the method for dealing with spatial gaps and uneven sampling, and the inclusion or exclusion of terrestrial, shallow and deep marine data, and the differential reporting of the results complicate comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust method for inferring temperature gradients that can be readily used with different proxy types and with very patchy records, and that fully quantifies the modeled temperature gradient with uncertainty. Applying this method to the EECO proxy record, we quantify the latitudinal SST gradient and the global mean SST.

@@

>>>>

---
title: "A Bayesian method for latitudinal temperature gradients in the early Eocene"

@@

<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->

@@

Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g. @Hollis2019], and global temperature reconstructions based on those records alone are prone to bias [@Jones2022]. Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps. A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. Although some more recent climate models have been able to better align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust and relatively simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

Additional model runs with simulated SST data show that latitudinal gradients following a flat, linear or quadratic function can also be adequately reconstructed by our model (Fig. S4). The generalised logistic function underlying the reconstruction leads to the inferred gradient being pushed towards a sigmoidal shape in the absence of proxies (see very high latitudes in Fig S4a,c). This is intended, as in the absence of data, the shape of the reconstructed gradient should be broadly similar to the sigmoidal shape of the modern SST gradient, as it is the only empirically observed SST gradient.
